mlr3spatiotempcv: Spatiotemporal Resampling Methods for 'mlr3

### 2020-11-15 joss/joss.Rmd
First edit of joss article

At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: (CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012]).
There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017, @pohjankukka2017]

The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather most available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.

@@

Data which includes spatial or temporal information requires special treatment in machine learning; similar to cost-sensitive, functional, multilabel, ordinal or survival datasets [@mlr3book].
In contrast to non-spatial/non-temporal data, observations inherit a natural grouping, either in space or time or in both space and time [@legendre1993].
This grouping causes observations to be autocorrelated, either in space (spatial autocorrelation (SAC)), time (temporal autocorrelation (TAC)) or both space and time (spatiotemporal autocorrelation (STAC)).
For simplicity, the acronym STAC is used as a generic term in the following chapter for all the different characteristics introduced above.

@@

This dataset contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
The landslides were mapped from aerial photos taken in 2000.
The dataset is well suited to serve as an example because it is relatively small and of course due to the spatial nature of the observations.
Please refer to @muenchow2012 for a detailed description of the dataset.

To account for the spatial autocorrelation probably present in the landslide data, we will make use one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
The grouping has the effect that train and test data are more separated in space as they would be by conducting a random partitioning, thereby reducing the effect of STAC.

By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the spatial CV approach.
However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.

Rephrased in other words: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
Results from a SptCV are not unbiased but more close in reflecting the actual performance of the learner on unknown datasets.

@@

Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
Remember that the better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.

@@

This example used an already created task via the sugar function `tsk()`.
In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.

>>>>

At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@brenning2012].
There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017, @pohjankukka2017]

The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simplified and generic interface.
The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework.
This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline.
The hope is that the availability of such a package encourages researchers to more often apply spatiotemporal resampling methods in scientific case studies.

@@

Data which includes spatial or temporal information requires special treatment in machine learning; similar to cost-sensitive, functional, multilabel, ordinal or survival datasets [@mlr3book].
In contrast to non-spatial/non-temporal data, observations inherit a natural grouping, either in space or time or in both space and time [@legendre1993].
This grouping causes observations to be autocorrelated, either in space (spatial autocorrelation (SAC)), time (temporal autocorrelation (TAC)) or both space and time (spatiotemporal autocorrelation (STAC)).
For simplicity, the acronym STAC is used as a generic term in the following discussion for all the different characteristics introduced above.

@@

This dataset contains information on the occurrence of landslides (binary) in the Andes of Southern Ecuador.
The landslides were mapped from aerial photos taken in 2000.
The dataset is well suited to serve as an example because it is relatively small and, of course, due to the spatial nature of the observations.
Please refer to @muenchow2012 for a detailed description of the dataset.

To account for the spatial autocorrelation probably present in the landslide data, we will make use of one of the most used spatial partitioning methods, a cluster-based $k$-means grouping [@brenning2012], (`"spcv_coords"` in [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com)).
This method performs a clustering in 2D space which contrasts with the commonly used random partitioning for non-spatial data.
The grouping has the effect that train and test data are more separated in space than they would be by conducting a random partitioning, thereby reducing the effect of STAC.

By contrast, when using the classical random partitioning approach with spatial data, train and test observations would be located side-by-side across the full study area (a visual example is provided further below).
This leads to a high similarity between train and test sets, resulting in "better" but over-optimistic performance estimates in every fold of a CV compared to the Spatial CV (SpCV) approach.
However, these low error rates are mainly caused due to the STAC in the observations and the lack of appropriate partitioning methods and not by the power of the fitted model.

Rephrased in other words: the higher performance results in a non-spatial CV scenario should not be relied on or used for reporting.
The positive bias in these results is substantial and the actual performance of the model is nowhere near this result.
Results from a SpCV are not unbiased but are closer in reflecting the actual performance of the learner on unknown datasets.

@@

Here, the estimated AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation.
Remember that the better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.

@@

This example used an already created task via the convenience function `tsk()`.
In practice however, one needs to create a spatiotemporal task via `TaskClassifST()`/`TaskRegrST()` and set the `crs` argument.
